---
title: "Use of For Loops"
author: "Daniel Schiff"
date: "`r format(Sys.Date(),'%B %d, %Y')`"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Importance of For Loops

In almost everything we do, we see repetition resulting in the development of
stable routines and production. For example, Henry Ford was able to find
reproducibility in the assembly line. Manufacturing car after car the same way
to improve cost, and efficiency. Like Ford's method, modern computers can repeat
a process over and over again with accuracy and exactness based on the
information (code) it is fed. In fact, this
is one of the great advantages of computing machines.

In R, we have a several functions that will do tasks repeatedly while some condition(s) are met.

+ **while loops**, which do something _while_ a statement or group of statements are true. 
+ **repeat loops**, which will continue to do something until it is told to
not continue. If it is never told not to continue,  it will keep running infinitely.
+ **for loops**, which do something while going through values of a sequence or list that is fed to it. We'll look at this loop in detail since it is so commonly used.

We develop an example of how to use a _for loop_ for calculating probabilities 
when sampling without
replacement. Suppose we have 100 Skittles, and we know that 3 would kill you if 
you ate them, while the other 97 are 100% safe.

The question we want to find out is, given you have eaten a certain number of
Skittles already, *what is the chance that you'll remain alive after eating the
next Skittle*.

The first time you eat one of the Skittles, you
have a 97 out of 100 chance of staying alive (97%). The odds of
staying alive will go down, as you eat more Skittles (sampling without
replacement). By the time you eat the 31st Skittle, the probability that the
next Skittle won't kill you is only 25%.

> In the jargon of probability theory, we are simulating a [**geometric distribution**](https://en.wikipedia.org/wiki/Geometric_distribution), 
which can be thought of as the probability distribution of the number of successes (staying alive) before experiencing a failure (dying)

```{r, warning=FALSE, message=FALSE}
require(tidyverse)
# Set p to NA
p <- NA
# Set a numerator
n <- 0
# Set Denominator
d <- 3
# Set an empty data frame
output <- NA

e <- 1
# expected life value, if you eat no skittles 100% chance of living or 1

# Write a for loop to produce probability that they will not kill you
for (i in 98:1) {
  # make the values into a vector
  out <- data.frame(i, n, d, p, e)
  # Add one to the numerator
  n <- n + 1
  # Add one to the denominator
  d <- d + 1
  # combine the vector into the data frame
  output <- rbind(output, out) %>%
    mutate(p = n / d, e = e * p) %>%
    drop_na(p) # drop if there are a null vector
}

# Include a Graphic

g <- ggplot(output, aes(x = i, y = e)) +
  geom_point(shape = 1, color = "red") +
  xlab("Number of Skittles Consumed") +
  ylab("Life Expectancy") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_bw() +
  labs(title = "Skittle Probability Rate")

g
```

## Lets look more into for loops 

_For loops_ are a very useful method to use for many repetitive tasks, since it 
automates the process and also, won't make a mistake `r emo::ji("happy")`. For 
example, if we need to create the same data visualization for multiple data sets,
we can use a for loop to create them without the hassle, and danger, of copying 
and pasting our code. The following is an example of drawing histograms for the 
distribution of a random variable, as we draw larger and larger samples from its
distribution. We loop over a sequence of values that specify the number of samples
we are drawing (`seq(25, 500, 25)`), and as we look see, the histograms slowly
start to look like the familiar bell curve. Notice that the `seq` function allows
us to create arbitrary regular sequences, so we're not restricted to 1,2,3,...



```{r, warning=FALSE, message=FALSE}

for (m in seq(50, 500, 50)) {

  # Set the seed for standard sampling
  set.seed(435)

  # Sample a random normal distribuition
  x <- rnorm(n = m, mean = 0, sd = 1)
  x <- data.frame(x)

  p <- ggplot(x, aes(x)) +
    geom_histogram(bins = 50) +
    labs(title = paste0("N = ", m))

  print(p)
}
```


## We Can Loop For Categorical or Factor Data

When looking at loops, at first numbers come to mind.
However, _for loops_ can be run on any arbitray **list** of objects. These could be 
a list of data.frame objects, a list of models, or a list of words. In the following loop, we look at the `species` variable in the iris data set. We do a complex loop of group_by() and summarize() as seen in the bit of code below it. The idea is to show 
how a list of words derived from your data can be used in a for loop. 


```{r}
outputiris <- list() # create an empty list

for (sp in levels(iris$Species)) {
  print(sp)
  # I am just kidding lets take it to the next step

  # Lets group by using a filter to filter the type of species we would want
  outiris <- filter(iris, Species == sp) %>%
    # Lets set the means
    summarise(
      Species = sp, slMean = mean(Sepal.Length), swMean = mean(Sepal.Width),
      plMean = mean(Petal.Length), pwMean = mean(Petal.Width)
    )
  outputiris[[sp]] <- outiris # Save the result into the list, which is now a named list
}

outputiris <- bind_rows(outputiris, .id='Species')
print(outputiris)
```

There is an easier and quicker way to do this, but the previous example was meant
to show how we _can_ do it using for loops.

```{r}
 iris %>%
  group_by(Species) %>%
  summarise(
    slMean = mean(Sepal.Length), swMean = mean(Sepal.Width),
    plMean = mean(Petal.Length), pwMean = mean(Petal.Width)
  )
```
